---
title: "Minneapolis"
author: "Eleanor Noble"
date: "12/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Install packages
```{r}
library(pacman)
p_load(tidyverse, tidycensus, janitor, sf, skimr, beepr, stringr, stringdist,pacman)
```
Read in data
```{r}
henn_parcels_raw<-read_csv("C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/Hennepin_County_Parcels_12.4.2021.csv")%>%janitor::clean_names()

mpls_rentals<-read_csv("C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/Active_Rental_Licenses_minneapolis_12.3.2021.csv")%>%janitor::clean_names()

zipco_lu <- read_csv("C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/zipco_lu.csv")
```

Clean columns
```{r}
#changing sale year to a date
henn_parcels <- henn_parcels_raw%>%
  mutate(sale_date = lubridate::ymd(sale_date))

#change all characters to lowercase
henn_parcels<-henn_parcels%>%
  mutate_if(.predicate = is.character, .funs = tolower)

henn_parcels<-henn_parcels%>%
  select(pid,
         house_no,
         street_nm,
         condo_no,
         mailing_munic_nm,
         zip_cd, 
         owner_nm, 
         taxpayer_nm,
         taxpayer_nm_1, 
         taxpayer_nm_2,
         taxpayer_nm_3,
         munic_nm,
         build_yr,
         sale_date,
         sale_price,
         sale_code,
         tax_tot,
         pr_typ_nm1,
         hmstd_cd1
         )

#change NAs to nothing in character columns
na_to_nothing <- function(x) ifelse(is.na(x),'',x)

#cleans the strings in character columns to be all uppercase, removes whitespace, and removes NAs
henn_parcels<-henn_parcels%>%
  mutate_if(.predicate = is.factor, .funs = as.character)%>%
  mutate_if(.predicate = is.character,.funs = na_to_nothing)%>%
  mutate_if(.predicate = is.character, .funs = trimws)%>%
  mutate_if(.predicate = is.character, .funs = tolower)

#clean up the parcel address columns to be simplified to just building number and streetname columns or full address
henn_parcels<-henn_parcels%>%
  mutate(address_full = paste(house_no, street_nm,condo_no, mailing_munic_nm, zip_cd, "MN"))%>%
  select(-street_nm,
         -condo_no)

#clean taxpayer address to get a taxpayer zipcode and taxpayer full address
henn_parcels<-henn_parcels%>%
  mutate(tax_address_full = paste(taxpayer_nm_1,taxpayer_nm_2,taxpayer_nm_3),
         tax_zip = regmatches(tax_address_full, gregexpr('[0-9]{5}(-[0-9]{4})?(?!.*[0-9]{5}(-[0-9]{4})?)', tax_address_full, perl = TRUE)))%>%
  select(-taxpayer_nm_1,
         -taxpayer_nm_2,
         -taxpayer_nm_3)
```
create a parcel and taxpayer spreadsheet for geocoding
```{r}
henn_parcels<-henn_parcels%>%
  mutate(id = row_number())

parcel_add<-henn_parcels%>%
  select(id,
         address_full)

tax_add<-henn_parcels%>%
  select(id,
         tax_address_full)

mpls_rental_owner<-mpls_rentals%>%
  mutate(owner_add_full = paste(owner_address1, owner_address2, owner_city, owner_state, owner_zip))%>%
  select(apn,
         owner_add_full)

write_csv(mpls_rental_owner, "C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/rental_owner_add_for_geocoding.12.15.2021.csv")
write_csv(parcel_add, "C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/henn_parcel_add_for_geocoding.12.15.2021.csv")
write_csv(tax_add, "C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/henn_tax_add_for_geocoding.12.15.2021.csv")
```

read in the geocoded addresses
```{r}
henn_parcels_geocoded<-read_csv("C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/henn_parcel_add_for_geocoding.12.15.2021-geocoded.csv", col_types = cols(.default = "c"))

henn_parcels_geocoded<-henn_parcels_geocoded%>%
  select(id,
         Match_addr,
         Longitude,
         Latitude,
         Postal)%>%
  rename(address_full_clean = Match_addr,
         parcel_zip = Postal,
         parcel_long = Longitude,
         parcel_lat = Latitude
  )

henn_tax_addr_geocoded<-read_csv("C:/Users/enoble/Box/Landlord Blog/Data/Minneapolis/henn_tax_add_for_geocoding.12.15.2021-geocoded.csv", col_types = cols(.default = "c"))

henn_tax_addr_geocoded<-henn_tax_addr_geocoded%>%
  select(id,
         Match_addr,
         Longitude,
         Latitude,
         Postal)%>%
    rename(tax_full_clean = Match_addr,
         tax_zip = Postal,
         tax_long=Longitude,
         tax_lat = Latitude
  )

geocoded<-henn_parcels_geocoded%>%
  left_join(henn_tax_addr_geocoded, by = c("id"="id"))

#change a few variable names to match previous code
henn_parcels<-henn_parcels%>%
  rename(use1_desc = pr_typ_nm1,
         homestead = hmstd_cd1)

#only keep needed columns and then join cleaned geocoded columns
henn_parcels<-henn_parcels%>%
  select(pid,
         zip_cd,
         munic_nm,
         build_yr,
         sale_date,
         sale_price,
         tax_tot,
         use1_desc,
         homestead,
         id)

henn_parcels<-henn_parcels%>%
  mutate(id = as.character(id))%>%
  left_join(geocoded, by = ("id"="id"))
```

join the mpls rentals to the parcels dataset
```{r}
mpls_rentals_henn_parcels<-henn_parcels%>%
  left_join(mpls_rentals, by = c("pid" = "apn"))
```
get the taxpayer and owner names back on there! oops

```{r}
tax_owner_names<-henn_parcels_raw%>%
  select(pid,
         owner_nm,
         taxpayer_nm)

mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  left_join(tax_owner_names, by = c("pid" = "pid"))
```

clean the columns again...
```{r}
#cleans the strings in character columns to be all uppercase, removes whitespace, and removes NAs
mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  mutate_if(.predicate = is.factor, .funs = as.character)%>%
  mutate_if(.predicate = is.character,.funs = na_to_nothing)%>%
  mutate_if(.predicate = is.character, .funs = trimws)%>%
  mutate_if(.predicate = is.character, .funs = tolower)

mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  rename(tax_address_full = tax_full_clean,
         address_full = address_full_clean)
```


```{r,zip codes for geography designations}
#Geographic look up table from Missouri Census Data Center
zipco_lu <- zipco_lu %>%
  mutate_all(as.character) %>%
  group_by(zcta5) %>%
  filter(afact == max(afact)) %>%
  ungroup() %>%
  mutate(stab = "MN")%>%
  select(tax_zip = zcta5, tax_state = stab, tax_co = cntyname)

# Metro County vector
metro_cnty <- c("Hennepin MN", "Ramsey MN", "Anoka MN", "Dakota MN", "Washington MN", "Scott MN", "Carver MN")
```

ownership classification
```{r, adding ownership classification}
### ownership class ###
llc <- "/\bllp\\b|\\blllp\\b|\\blllc\\b|\\blimited\\b|\\bllc|\\blp"

coop <- "coop|cooperative"

trust<- "trtee|trustee|trust|trst|\\breit\\b"

private_ed  <- "k-12|academies|college|higher ed|private academies|univ of minn|u of m|university of minnesota|university|school|academy|college|school"

public <- "minneapois|airport|transitional housing|federal|county|public|municiple|mpha|state property|transit|council|mpha|municiapl|
municipal|public hospital|state property|mpls|library|right of way|row|public works|hennepin forfeited|city of|federal|fortified|henn co|henn cty|minnesota historical|municipal|hra|board|ramsey|council|metropolitan"

nonprof <- "bptst chrc|islamic|non profit|charitable|muslim|synagogue|mosque|non profit|catholic|baptist|methodist|jewish|urban homeworks|community hsg|com hsng|land bank|charity|lutheran"

corp <- "\\bcorp\\b|\\bbankers\\b|\\bcorporation\\b|\\bagency\\b|\\breal estate\\b|\\bmgmt\\b|\\basset\\b|\\bmanagement\\b|\\bbank\\b|\\binc\\b|\\bcompany\\b|\\binvest\\b|\\binvestment\\b|\\binvestments\\b|\\bservicing\\b|\\bmtg\\b|federal natl mortgage assn"

utility<- "railroad|power co|rr co|henn co|railway|railroad property|railroad property"

#create fields for if the parcel and tax address are the same, single family
mpls_rentals_henn_parcels <- mpls_rentals_henn_parcels %>% mutate(

    same_address = ifelse(tax_address_full==address_full, 1, 0),

    sf = ifelse(use1_desc %in% c("res 1 unit", "residential"), 1, 0), 

    use1_desc = ifelse(use1_desc %in% c("vacant land - residential", "res v land", "vacant land-apartment"), "vacant", use1_desc),

    sale_year = lubridate::year(sale_date)

  ) %>%
  left_join(zipco_lu, by = "tax_zip")

# skim(henn_parcels)

######## 
#this creates categories for if the taxpayer and owner are at the same address, theres a PO box (a lot of these so we need to filter them out eventually), if they're in the same zipcode, the same region (metro region), or out of the metro region. The sum class is a test to see if there are multiple location relationship hits for a single parcel
mpls_rentals_henn_parcels  <- mpls_rentals_henn_parcels  %>%
  mutate(

    homestead = ifelse(homestead == "f", "p", homestead),

    g_sadd = case_when(
      same_address == 1 ~ 1,
      TRUE               ~ 0),

    g_box  = case_when(
      (str_detect(tax_address_full,"box")) & same_address == 0 ~ 1,
      TRUE                                                         ~ 0
    ),

    g_szip   = case_when(
      zip_cd == tax_zip & g_sadd == 0 ~ 1,
      TRUE                          ~ 0),

    g_sreg  = case_when(
      tax_zip %in% metro_cnty & g_szip == 0 & g_sadd == 0  ~ 1,
      TRUE                                                 ~ 0),

    g_oreg = case_when(
      g_sreg == 0 & g_szip == 0 & g_sadd == 0  ~ 1,
     TRUE                                                 ~ 0),

    sumclass = g_sadd + g_szip 
    + g_sreg + g_oreg,

    owner_geo = case_when(
      g_sadd == 1  ~ "sadd",
      g_box  == 1  ~ "sbox",
      g_szip == 1  ~ "szip",
      g_sreg == 1  ~ "sreg",
      g_oreg == 1  ~ "oreg"
    ))

# par geo index
# Geography lu
#####
#this is the actual ownership classification. Owner type and owner typed are just different ways of grouping the categories 
mpls_rentals_henn_parcels <- mpls_rentals_henn_parcels %>%
  mutate(
    llc  = case_when(
      (str_detect(owner_name, llc) | str_detect(applicant_name, llc)
       |str_detect(owner_name, llc) | str_detect(taxpayer_nm, llc)) & tax_tot>0 ~ 1,
      TRUE                                                              ~0),

    corp =
      case_when(
        (str_detect(taxpayer_nm, corp)|str_detect(applicant_name, corp)|str_detect(owner_name, corp))
        & tax_tot>0 & homestead == "n" & llc == 0 ~ 1,
        TRUE                                                               ~ 0),

    np   = case_when(
      tax_tot==0                                          ~ 1,
      (str_detect(applicant_name, nonprof)|str_detect(taxpayer_nm, nonprof))  ~ 1,
      TRUE                                                         ~ 0
    ),

    gov  = case_when(
      (str_detect(taxpayer_nm, public)| str_detect(taxpayer_nm, private_ed) | str_detect(taxpayer_nm, utility))&
        homestead == "n" & llc == 0 & corp == 0 & tax_tot==0 ~ 1,
      TRUE                                                         ~ 0
    ),


    hmstd = case_when(
      (homestead == "h" | homestead == "p")  ~ 1,
      TRUE             ~ 0
    ),

    own_sum = llc + corp + gov + np + hmstd,

    owner_typed = case_when(
      np      == 1  ~ "np",
      llc     == 1 & np == 0  ~ "llc",
      corp    == 1 & np == 0  ~ "corp",
      gov     == 1  ~ "gov",
      own_sum == 0  ~ "othpriv",
      hmstd   == 1  ~ "hmstd"),

    owner_type = case_when(
      gov == 1 | np ==  1 ~ "govnp",
      llc == 1 | corp == 1 ~ "corpllc",
      hmstd   == 1  ~ "hmstd",
      own_sum == 0  ~ "othpriv"))
  # select(-tax_more) %>%
  # distinct(pin, dataset, .keep_all = TRUE) 
# %>%
#   mutate(acres = sq_meters*0.000247105)

table(mpls_rentals_henn_parcels$owner_type)
```

Create a potential/likely rental category in the hennepin parcel dataset
```{r}
#this uses land use codes that were confirmed in the city wide dataset to have rental properties. I'm only including ones that have 20 or more known units to be sure, on the conservative side 

#land use that doesnt need a homestead to be considered a rental
rental_no_hmst<-"apartment|low income rental|cooperative housing|residential-two unit|triplex"
#land use that needs homestead
rental_hmst<-"residential|residential-zero lot line|townhouse|condominium"

  #this is the basic function for estimating if a parcel is a rental or not in ramsey county, it considers the use type, homestead, and number of units because the units data is good in ramsey
mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  mutate(
    rental  = case_when(
      ((str_detect(use1_desc, rental_no_hmst)&owner_type!="govnp")| #it's a definite rental in use1desc and not a non profit and not tax exempt
      (str_detect(use1_desc, rental_hmst)&owner_type!="govnp"&homestead=="n")) ~ 1,
      TRUE                                                         ~ 0
    ))

table(mpls_rentals_henn_parcels$rental) #this shows that there are 79735 rental properties in hennepin county, this seems right and is lower than units because its the parcels not units

mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  mutate(mpls_rental = ifelse(!is.na(objectid),1,0))%>%
  filter(mpls_rental==1|rental==1)

```

```{r, minneapolis rental license proxy test}
#only including mpls 2020 parcels
just_mpls_parcels<-mpls_rentals_henn_parcels%>%
  rename(city = munic_nm)%>%
  filter(city == "minneapolis")


#creating a column that shows correct rental classifications, false positives, false negatives, and correct negatives
just_mpls_parcels_rental_join<-just_mpls_parcels%>%
  mutate(correct_rental = ifelse(rental==TRUE&!is.na(objectid), "match","correct negative"),
         correct_rental = ifelse(rental==TRUE&is.na(objectid), "false positive",correct_rental),
         correct_rental = ifelse(rental==FALSE&!is.na(objectid), "false negative",correct_rental))

table(just_mpls_parcels_rental_join$correct_rental)

#here you can see that the homestead status variable didn't work in the initial rental proxy column
false_positives<-just_mpls_parcels_rental_join%>%
  filter(correct_rental=="false positive")
```

Create property owner size column
```{r}
#there are some points that won't join so we'll add them on to the data set
owner_size<-mpls_rentals_henn_parcels%>%
 dplyr::select(pid,tax_address_full,owner_nm,taxpayer_nm,licensed_units)%>%
  mutate(units = ifelse(is.na(licensed_units),1,licensed_units))

owner_size<-owner_size%>%
  mutate(tax_address_full = ifelse(is.na(tax_address_full)|tax_address_full=="","no tax add",tax_address_full),
         owner_nm = ifelse(is.na(owner_nm)|owner_nm=="", "no owner name", owner_nm),
         taxpayer_nm = ifelse(is.na(taxpayer_nm)|taxpayer_nm =="", "no tax name", taxpayer_nm))%>%
  group_by(tax_address_full)%>%
  mutate(tot_props_owned_by_taxpayer_add = sum(n()))%>%
  group_by(tax_address_full)%>%         
  mutate(tot_units_owned_by_taxpayer_add = sum(units))%>%
  group_by(owner_nm)%>%
  mutate(tot_props_owned_by_owner_name = sum(n()),
         tot_units_owned_by_owner_name = sum(units))%>%
  group_by(taxpayer_nm)%>%
  mutate(tot_props_owned_by_taxpayer_nm = sum(n()),
         tot_units_owned_by_taxpayer_nm = sum(units))

owner_size_save<-owner_size

owner_size<-owner_size_save

owner_size<-owner_size%>%
  mutate(max_units_by_owner = max(tot_units_owned_by_taxpayer_add,tot_units_owned_by_owner_name,tot_units_owned_by_taxpayer_nm),
         max_props_by_owner = max(tot_props_owned_by_owner_name, tot_props_owned_by_taxpayer_add, tot_props_owned_by_taxpayer_nm))%>%
  select(pid,
         max_units_by_owner,
         max_props_by_owner
         )
#now we join this back onto the points data set

mpls_rentals_henn_parcels<-mpls_rentals_henn_parcels%>%
  left_join(owner_size, by = c("pid"="pid"))

mpls_rentals<-mpls_rentals_henn_parcels%>%
  filter(!is.na(objectid))
#create data frame that just has the object IDs of the licensed rentals from mpls and do a filtering join
mpls_rentals<-unique(mpls_rentals)

```
only choose needed columns
```{r}
mpls_rentals<-mpls_rentals%>%
  select(pid,
         zip_cd,
         munic_nm,
         use1_desc,
         address_full,
         tax_address_full,
         tax_lat,
         tax_long,
         tax_zip,
         x,
         y,
         owner_state,
         owner_geo,
         owner_type,
         max_units_by_owner,
         max_props_by_owner,
         licensed_units,
         owner_name)
```

RCAA/RCAP
```{r}
##what are the demographics of the areas they live in
acs_19_vars <- load_variables(2019, "acs5", cache = TRUE)

my_vars<- c(pop = "B01003_001",
            pov_total = "B17001_001",
            in_pov = "B17001_002",
            med_hh_inc="B19013_001",
            white_alone = "B03002_003", 
            latino = "B03003_003",
            black = "B03002_004", 
            asian = "B03002_006", 
            native="B03002_005",
            api="B03002_007",
            occ_status="B25002_001",
            vacant="B25002_003",
            med_year_built="B25035_001",
            total_tenure="B25003_001",
            owners="B25003_002",
            renters="B25003_003")

MN_demos <- get_acs(geography = "tract", variables = my_vars,
        state="MN",geometry = TRUE) %>%
  select(-moe) %>%
  spread(variable, estimate) %>%
  filter(pop!=0) %>%
  mutate(pct_renter = (renters/total_tenure),
         pct_owner = (owners/total_tenure),
         pct_pov=(in_pov/pov_total),
         pov_rate = (in_pov/pov_total),
         pct_not_white = ((pop-white_alone)/pop),
         pct_white = (white_alone/pop),
         pct_latino = (latino/pop),
         pct_black = (black/pop),
         pct_asian = (asian/pop)) %>%
  st_transform(26792)

henn_tract_demos <- MN_demos %>%
  filter(grepl('Hennepin', NAME)) %>%
  mutate(
    #40+ % pov and over 50% not white
    RCAP = ifelse(pct_pov > .4 & pct_not_white > .5, 1,0),
    ###Over 80% white and over 2x national median inc, which is 65889
    RCAA = ifelse(pct_not_white <= .2 & med_hh_inc >=131778, 1, 0)
    )

mpls_rentals_sf <- st_as_sf(mpls_rentals, coords = c("x", "y"),crs = 4326, agr = "constant")%>%
  st_transform(crs = 26792)

mpls_owners_sf <- st_as_sf(mpls_rentals, coords = c("tax_long", "tax_lat"),crs = 4326, agr = "constant")%>%
  st_transform(crs = 26792)

#spatial join the rentals to the tract demographics
rentals_demos <- mpls_rentals_sf %>%
  st_transform(26792) %>%
  st_join(henn_tract_demos, join=st_intersects)

# spatial join the landlords to the tract
landlords_demos <- mpls_owners_sf %>%
  st_transform(26792) %>%
  st_join(henn_tract_demos, join=st_intersects)

#map it to see rcaas and rcaps
rcaa_map<-henn_tract_demos %>%
  ggplot() +
  geom_sf(mapping = aes(fill = factor(RCAA))) +
  scale_fill_discrete() +
  labs(fill = "RCAA")


#there are 95818 rental units in this data based on the licensed units

#overall landlord count
landlords_count<-landlords_demos%>%
  select(tax_address_full)

#there are 11131 unique landlords in hennepin county
landlords_in_henn<-landlords_demos%>%
  filter(!is.na(RCAP))

sum(landlords_in_henn$licensed_units)

LLC_owners<-landlords_demos%>%
  filter(owner_type=="corpllc")%>%
  select(tax_address_full)

LLC_owners<-unique(LLC_owners)

sum(LLC_owners$licensed_units)

#mom and pop is if they live in the county and have fewer than 10 units
mom_and_pop<-landlords_demos%>%
  filter(!is.na(RCAP)&
           (max_units_by_owner<=3))%>%
  select(tax_address_full)

mom_and_pop<-unique(mom_and_pop)

sum(mom_and_pop$licensed_units)
```
Looking at property portfolio distribution
```{r}
boxplot(landlords_demos$max_props_by_owner)

ggplot(landlords_demos, aes(x=max_props_by_owner)) + 
    geom_histogram(aes(y=..density..),      # Histogram with density instead of count on y-axis
                   binwidth=.5,
                   colour="black", fill="white") +
    geom_density(alpha=.2, fill="#FF6666")  # Overlay with transparent density plot

properties_dist<-landlords_demos$max_props_by_owner

quantile(properties_dist)
```

RCAA/RCAP analysis
```{r}
#overall mpls RCAA/RCAP makeup

landlords_rcaaps<-landlords_demos%>%
  rename(rcaa_ll = RCAA,
         rcap_ll = RCAP)%>%
  select(pid,
         rcaa_ll,
         rcap_ll)

landlords_rcaaps<-unique(landlords_rcaaps)

rentals_rcaaps<-rentals_demos%>%
  rename(rcaa_rental = RCAA,
         rcap_rental = RCAP)%>%
  select(rcaa_rental,
         rcap_rental,
         pid)

rentals_rcaaps<-unique(rentals_rcaaps)

mpls_rcaa_rcap<-mpls_rentals%>%
  select(pid,
         tax_address_full,
         licensed_units,
         max_units_by_owner,
         max_props_by_owner)%>%
  left_join(landlords_rcaaps, by = c("pid" = "pid"))%>%
  left_join(rentals_rcaaps, by = c("pid" = "pid"))

mpls_rcaa_rcap<-mpls_rcaa_rcap%>%
  group_by(rcaa_rental,
           rcap_rental,
           rcaa_ll,
           rcap_ll,
           tax_address_full)%>%
  mutate(ll_count = n())
  group_by(rcaa_rental,
           rcap_rental,
           rcaa_ll,
           rcap_ll)%>%
  summarise(props = n(),
            units = sum(licensed_units))
  
#mom and pop RCAA/RCAP makeup
# mpls_rcaa_rcap<-mpls_rcaa_rcap%>%
#   group_by(rcaa_ll, rcap_ll,rcaa_rental,rcap_rental)%>%
#   mutate()
```

